---
title: "TP53 CDS plot"
author: "Jip van Dinter"
date: "12/2/2021"
output: html_document
---

This script requires the bed files that are output from riboseqc. Please make sure these files are in the data_dir.

```{r}
data_dir <- folder with coverage tracks from RiboseQC
save_dir <- save dir
color_scheme <- c("#8F8F93","#43C1F0","#FF8800")
```

```{r libraries}
library(ggplot2)
library(rtracklayer)
library(scales)
```

## Load the data

The position DF is used to connect CDS positions to coverage in the later script chunks.

```{r Position DF}
single_nt_tp53_gr <- Granges(read.table(paste(data_dir,"tp53_single_nt_windows.txt", sep = "/"), 
                                        header = T, sep = "\t"))
rel_pos_df <- read.table(paste(data_dir,"tp53_cds_position.txt", sep = "/"), 
                         header = T, sep = "\t")

```

Load in the Read Coverage data from RiboseQC, only using - strand as TP53 is located on the - strand.

```{r RiboseQC Unique reads}
ribo_cov_dfs_uniq <- list()
ribo_cov_list <- list.files(path = data_dir, pattern = "*coverage_uniq_minus*",recursive = T) 

for (i in seq_along(ribo_cov_list)) {

  sample_file = ribo_cov_list[i]
  sample_id = sub(".*H1299-","",sample_file)
  sample_id = sub("_coverage.*","",sample_id)
  print(paste("Parsing", sample_id, sep = " "))
  f = import.bed(paste(data_dir,sample_file, sep = "/"))
  strand(f) <- "-"
  colnames(mcols(f)) <- "coverage"
  
  # Calculate P site coverage over TP53 CDS region
  exact_hits <- findOverlaps(query = single_nt_tp53_gr, subject = f, type = "equal")
      
  query_exact_hits <- single_nt_tp53_gr[queryHits(exact_hits)]
  subject_exact_hits <- f[subjectHits(exact_hits)]
  ribo_cov_df <- as.data.frame(subject_exact_hits)
  
  # Annotate DF with sample data
  ribo_cov_df$sample_id <- rep(sample_id,nrow(ribo_cov_df))
  ribo_cov_df$rel_pos <- rel_pos_df[which(rel_pos_df$pos %in% ribo_cov_df$start),]$rel_pos
  ribo_cov_df$vector <- ifelse(grepl("EV", ribo_cov_df$sample_id),"EV","R213X")
  ribo_cov_df$treatment <- ifelse(grepl("G418", ribo_cov_df$sample_id),"G418",
                        ifelse(grepl("5FUR",ribo_cov_df$sample_id),"5FUR","NT"))
  ribo_cov_df$treatment <- factor(ribo_cov_df$treatment, levels = c("NT","5FUR","G418"))
  ribo_cov_dfs_uniq[[sample_id]] <- ribo_cov_df
}

ribo_cov_df_uniq <- do.call("rbind",ribo_cov_dfs_uniq)
ribo_cov_df_uniq$coverage <- as.numeric(ribo_cov_df_uniq$coverage)
```

Create DF for coverage difference between Untreated and Treated samples

```{r Remove background reads}
ribo_cov_delta_dfs <- list()
ribo_cov_delta_list <- list.files(path = data_dir, pattern = "*coverage_uniq_minus*",recursive = T) 

for (i in seq_along(ribo_cov_delta_list)) {

  sample_file = ribo_cov_delta_list[i]
  sample_id = sub(".*H1299-","",sample_file)
  sample_id = sub("_coverage.*","",sample_id)
  print(paste("Parsing", sample_id, sep = " "))
  f = import.bed(paste(data_dir,sample_file, sep = "/"))
  strand(f) <- "-"
  colnames(mcols(f)) <- "coverage"
  
  # Calculate P site coverage over TP53 CDS region
  exact_hits <- findOverlaps(query = single_nt_tp53_gr, subject = f, type = "equal")
      
  query_exact_hits <- single_nt_tp53_gr[queryHits(exact_hits)]
  subject_exact_hits <- f[subjectHits(exact_hits)]
  ribo_cov_delta_df <- as.data.frame(subject_exact_hits)
  
  # Annotate DF with sample data
  ribo_cov_delta_df$sample_id <- rep(sample_id,nrow(ribo_cov_delta_df))
  ribo_cov_delta_df$rel_pos <- rel_pos_df[which(rel_pos_df$pos %in% ribo_cov_delta_df$start),]$rel_pos
  
  df <- left_join(rel_pos_df, ribo_cov_delta_df[,c("rel_pos","coverage")])
  df[is.na(df)] <- 0
  colnames(df) <- c("pos","rel_pos",sample_id)
  ribo_cov_delta_dfs[[sample_id]] <- df[,sample_id]
}

ribo_cov_delta_df <- as.data.frame(do.call("cbind",ribo_cov_delta_dfs))
ribo_cov_delta_df <- bind_rows(lapply(ribo_cov_delta_df,as.numeric))
ribo_cov_ev <- ribo_cov_delta_df[,1:12]
ribo_cov_r213x <- ribo_cov_delta_df[,13:24]

# Calculate sums for delta 
ev_nt <- rowSums(ribo_cov_ev[1:4])
ev_g418 <- rowSums(ribo_cov_ev[5:8])
ev_5fur <- rowSums(ribo_cov_ev[9:12])
r213x_nt <- rowSums(ribo_cov_r213x[1:4])
r213x_g418 <- rowSums(ribo_cov_r213x[5:8])
r213x_5fur <- rowSums(ribo_cov_r213x[9:12])

# Collect in one table
plot_delta <- data.frame(rel_pos = rel_pos_df$rel_pos,
                         "EV_5FUR" = ev_5fur - ev_nt,
                         "EV_G418" = ev_g418 - ev_nt,
                         "R213X_5FUR" = r213x_5fur - r213x_nt,
                         "R213X_G418" = r213x_g418 - r213x_nt) %>%
  pivot_longer(!rel_pos, names_to = "group_id", values_to = "coverage")

plot_delta$vector <- ifelse(grepl("EV", plot_delta$group_id),"EV","R213X")
plot_delta$treatment <- ifelse(grepl("G418", plot_delta$group_id),"G418","5FUR")
plot_delta$treatment <- factor(plot_delta$treatment, levels = c("NT","5FUR","G418"))
```

Explanation of chosen numbers:

### CDS
1:636
### R213X codon
637
638
639
### CDS
length(640:nrow(rel_pos))

Coverage data for ratios
```{r RiboseQC ratio}
ribo_ratio_dfs <- list()
ribo_ratio_list <- list.files(path = data_dir, pattern = ".*R213X.*coverage_uniq_minus.bedgraph", recursive = T) 

for (i in seq_along(ribo_ratio_list)) {

  sample_file = ribo_ratio_list[i]
  sample_id = sub(".*H1299-","",sample_file)
  sample_id = sub("_coverage.*","",sample_id)
  print(paste("Parsing", sample_id, sep = " "))
  f = import.bed(paste(data_dir,sample_file, sep = "/"))
  strand(f) <- "-"
  colnames(mcols(f)) <- "coverage"
  
  # Calculate P site coverage over TP53 CDS region
  exact_hits <- findOverlaps(query = single_nt_tp53_gr, subject = f, type = "equal")
      
  query_exact_hits <- single_nt_tp53_gr[queryHits(exact_hits)]
  subject_exact_hits <- f[subjectHits(exact_hits)]
  ribo_ratio_df <- as.data.frame(subject_exact_hits)
  ribo_ratio_df$coverage <- as.numeric(ribo_ratio_df$coverage)
  ribo_ratio_df$rel_pos <- rel_pos_df[which(rel_pos_df$pos %in% ribo_ratio_df$start),]$rel_pos
  ribo_ratio_df$frame <- as.character(ribo_ratio_df$rel_pos %% 3)
  
  ribo_ratio_df_low <- ribo_ratio_df[which(ribo_ratio_df$rel_pos %in% 1:636),]
  ribo_ratio_df_high <- ribo_ratio_df[which(ribo_ratio_df$rel_pos %in% 640:nrow(rel_pos_df)),]
  ribo_ratio_sum_low <- sum(ribo_ratio_df_low$coverage)
  ribo_ratio_sum_high <- sum(ribo_ratio_df_high$coverage)
  treatment <- ifelse(grepl("G418", sample_id),"G418",
                        ifelse(grepl("5FUR",sample_id),"5FUR","NT"))
  treatment <- factor(treatment, levels = c("NT","5FUR","G418"))
  # Annotate DF with sample data
  ribo_ratio_df_new <- data.frame(sample_id = sample_id,
                             pre_stop_sum = ribo_ratio_sum_low,
                             post_stop_sum = ribo_ratio_sum_high,
                             pre_stop_sum_norm = ribo_ratio_sum_low / (636 / 3),
                             post_stop_sum_norm = ribo_ratio_sum_high / (length(640:nrow(rel_pos_df)) / 3),
                             ribo_ratio = (ribo_ratio_sum_high / (length(640:nrow(rel_pos_df)) / 3)) / (ribo_ratio_sum_low / (636 / 3)),
                             treatment = treatment
                             )
  ribo_ratio_dfs[[sample_id]] <- ribo_ratio_df_new
}

ribo_ratio_df <- do.call("rbind",ribo_ratio_dfs)
split_by_treatment <- split(ribo_ratio_df, ribo_ratio_df$treatment)

write.table(ribo_ratio_df, file = paste(save_dir,"ribo_tp53_ratio.txt",sep = "/"),
            quote = F, sep = ";", row.names = F, dec = ",")

```

Plot all the data:
```{r RiboseQC cov plot}
ggplot(data = ribo_cov_df_uniq, aes(x = rel_pos, xend = rel_pos, y = 0, yend = coverage)) +
  geom_segment(aes(col = treatment)) +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC R213X cov plot}
ggplot(data = ribo_cov_df_uniq[which(ribo_cov_df_uniq$vector == "R213X"),], aes(x = rel_pos, xend = rel_pos, y = 0, yend = coverage)) +
  geom_segment(aes(col = treatment)) +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC cov r213x zoom}
# Around R213X
ggplot(data = ribo_cov_df_uniq[which(ribo_cov_df_uniq$vector == "R213X"),], aes(x = rel_pos, xend = rel_pos, y = 0, yend = coverage)) +
  geom_segment(aes(col = treatment, group = sample_id)) +
  xlab("Relative CDS Position") +
  ylab("Normalised in-frame Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(637-100,637,637+150),
                     labels = c("","R213X",""),
                     limits = c(637-100,637+150)) +
  theme_bw() +
  facet_grid(~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_zoom.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC delta cov r213x plot}
ggplot(data = plot_delta[which(plot_delta$vector == "R213X"),], aes(x = rel_pos, 
                              xend = rel_pos,
                              y = 0, yend = coverage,
                              col = treatment)) +
  geom_segment() +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme[2:3]) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_delta.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC delta cov r213x zoom}
ggplot(data = plot_delta[which(plot_delta$vector == "R213X"),], 
       aes(x = rel_pos, xend = rel_pos,
           y = 0, yend = coverage,
           col = treatment)) +
  geom_segment() +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme[2:3]) +
  scale_x_continuous(limits = c(637,1179),
                     breaks = c(637,1179),
                     labels = c("R213X","stop")) +
  scale_y_continuous(limits = c(-2,5)) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_delta_zoom.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r Plot ratio}
# Errorbars with SE
ggplot(data = ribo_ratio_plotdata,aes(x = treatment, y = ratio)) +
  geom_boxplot(aes(fill = treatment)) +
  ggbeeswarm::geom_beeswarm() +
  scale_fill_manual(values = color_scheme) +
  xlab("Treatment Condition") +
  ylab("RPF Readthrough Ratio") +
  scale_y_continuous(labels = percent) +
  theme_classic()

ggsave(device = "pdf",
       filename = "readthrough_ratio_p53.pdf",
       width = 8,
       height = 8,
       path = save_dir)
```

Same plots, but with geom_line instead of geom_segment
```{r RiboseQC cov plot}
ggplot(data = ribo_cov_df_uniq, aes(x = rel_pos, y = coverage)) +
  geom_line(aes(col = treatment)) +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_line.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC R213X cov plot}
ggplot(data = ribo_cov_df_uniq[which(ribo_cov_df_uniq$vector == "R213X"),], aes(x = rel_pos, y = coverage)) +
  geom_line(aes(col = treatment)) +
  xlab("Relative CDS Position") +
  ylab("Normalised Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_line.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC cov r213x zoom}
# Around R213X
ggplot(data = ribo_cov_df_uniq[which(ribo_cov_df_uniq$vector == "R213X"),], aes(x = rel_pos, xend = rel_pos, y = 0, yend = coverage)) +
  geom_segment(aes(col = treatment, group = sample_id)) +
  xlab("Relative CDS Position") +
  ylab("Normalised in-frame Read Coverage") +
  scale_color_manual(values = color_scheme) +
  scale_x_continuous(breaks = c(637-100,637,637+150),
                     labels = c("","R213X",""),
                     limits = c(637-100,637+150)) +
  theme_bw() +
  facet_grid(~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_zoom_line.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC delta cov plot}
ggplot(data = plot_delta, aes(x = rel_pos, 
                              y =  coverage,
                              col = treatment)) +
  geom_line() +
  xlab("Relative CDS Position") +
  ylab("Difference in Norm. Coverage") +
  scale_color_manual(values = color_scheme[2:3]) +
  scale_x_continuous(breaks = c(0,637,1179),
                     labels = c("start","R213X","stop")) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_delta_line.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```

```{r RiboseQC delta cov r213x zoom}
plot_delta_line <- plot_delta
plot_delta_line[plot_delta_line == 0] <- NA
  
ggplot(data = plot_delta_line[which(plot_delta_line$vector == "R213X" & complete.cases(plot_delta_line)),], 
              aes(x = rel_pos,
                  y = coverage,
                  col = treatment)) +
  geom_line() +
  xlab("Relative CDS Position") +
  ylab("Difference in Norm. Coverage") +
  scale_color_manual(values = color_scheme[2:3]) +
  scale_x_continuous(limits = c(637,1179),
                     breaks = c(637,1179),
                     labels = c("R213X","stop")) +
  scale_y_continuous(limits = c(-2,5)) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggplot(data = plot_delta[which(plot_delta$vector == "R213X"),], 
              aes(x = rel_pos,
                  y = coverage,
                  col = treatment)) +
  geom_line() +
  xlab("Relative CDS Position") +
  ylab("Difference in Norm. Coverage") +
  scale_color_manual(values = color_scheme[2:3]) +
  scale_x_continuous(limits = c(637,1179),
                     breaks = c(637,1179),
                     labels = c("R213X","stop")) +
  scale_y_continuous(limits = c(-2,5)) +
  theme_bw() +
  facet_grid(vector ~ treatment)

ggsave(device = "pdf",
       filename = "tp53_cds_r213x_delta_zoom_line.pdf",
       width = 12,
       height = 8,
       path = save_dir)
```